Exploring probability distributions for bivariate temporal granularities

### 2019-11-24 main.Rmd
cricket story added

The application is not only restricted to temporal data. We provide an example of cricket to illustrate how this can be generalized in other applications. The Indian Premier League (IPL) is a professional Twenty20 cricket league in India contested by eight teams representing eight different cities in India. With eight teams, each team plays each other twice in a home-and-away round-robin format in the league phase. In a Twenty20 game the two teams have a single innings each, which is restricted to a maximum of 20 overs. Hence, in this format of cricket, a match will consist of 2 innings, an innings will consist of 20 overs, an over will consist of 6 balls. A hierarchy like table \ref{tab:hierarchy} can be construed for this game format.

@@

```{r hierarchy, echo=FALSE}
hierarchy_model <- tibble::tibble(units = c("ball", "over", "inning", "match"), 
                                  convert_fct = c(6, 20, 2, 1))
knitr::kable(hierarchy_model, caption = " A hierarchy table for T20 cricket")
```

```{r readdata}
library(gravitas)
library(tibble)
glimpse(cricket)
```

@@

```{r hierarchy2}

hierarchy_model <- tibble::tibble(
  units = c("index", "over", "inning", "match"),
  convert_fct = c(1, 20, 2, 1))

@@

First, we look at the distribution of runs per over across over of the innings and seasons in \ref{fig:seas-over-inning}. The distribution of runs per over has not significantly changed from 2008 to 2016. There is no clear pattern/trend that runs per over is increasing or decreasing across seasons. Hence, we work with subsets of seasons and try to see how the strategies of the winning teams differ across seasons or how in a particular season the strategy was different for the winning team and the ones who did not qualify for the playoffs.

Mumbai Indians(MI) and Chennai Super kings(CSK) are considered one of the best teams in IPL with multiple winning titles and always appearing in final 4 from 2010 to 2015. It would be interesting to observe their strategies throughout all matches in the two seasons. The following two questions might help us partially understand their strategies.

@@

Q2: How number of wickets and dot balls in the previous over affect the runs in the subsequent over across different overs of the innings?

@@

```{r seas-over-inning, fig.cap = "Quantile plot of runs per over across overs of different seasons."}
  cricket_tsibble_all %>% 
  prob_plot("over", 
            "season",
            hierarchy_model,
            response = "runs_per_over",
            plot_type = "quantile",
            quantile_prob = c(0.25, 0.5, 0.75), 
            symmetric = FALSE)  +
     theme( # remove the vertical grid lines
            panel.grid.major.x = element_blank() ,
            # explicitly set the horizontal lines (or they will disappear too)
            panel.grid.major.y = element_blank()) + scale_x_discrete(breaks = seq(2008, 2016, 3))
```

@@

cricket_tsibble %>%
  filter(batting_team %in% c("Mumbai Indians",
                             "Chennai Super Kings"))%>%
  prob_plot("inning", 
            "over",
  response = "runs_per_over",
  hierarchy_model,
  plot_type = "lv")
```

@@

```{r lv, fig.cap = "Letter value plot of run rate across overs of an inning and innings of the match for (top) Mumbai Indians and (bottom) Chennai Super Kings."}
# 
# 
# 
# MI <- cricket_tsibble_all %>%
#   filter(season %in% c(2010:2015), batting_team=="Mumbai Indians")
# 
# 
# CSK <- cricket_tsibble_all %>%
#   filter(season %in% c(2010:2015),
#          batting_team=="Kolkata Knight Riders",
#          inning %in% c(1,2))
# 
# 
#  MI %>%
#    gran_obs("inning", "over", hierarchy_model) %>%
#    knitr::kable(format = "markdown", caption = "Cross tabulation of Innings and Overs for Mumbai Indians: 2010-2015")
# 
#  CSK %>%
#    gran_obs("inning", "over", hierarchy_model) %>%
#    knitr::kable(format = "markdown", caption = "Cross tabulation of Innings and Overs for Mumbai Indians: 2010-2015")
# 
# MI %>%
#   prob_plot("inning", "over",
#                                hierarchy_model,
#                                response = "run_rate",
#                                plot_type = "lv",
#                                quantile_prob = c(0.1, 0.25, 0.5, 0.75, 0.9))  + ggtitle("")
# 
# 
# CSK %>%
#   prob_plot("inning", "over",
#                                hierarchy_model,
#                                response = "run_rate",
#                                plot_type = "lv",
#                                quantile_prob = c(0.1, 0.25, 0.5, 0.75, 0.9)) + ggtitle("")
# 
# cricket_tsibble_all %>% distinct(batting_team)

Q3: Is run rate set to reduce in subsequent over if fielding/bowling is good in the previous over? Between fielding and bowling which penalizes runs in the subsequent over more?

For establishing that the fielder fielded well in a particular over, we can see how many catches and run outs were made in that particular over. If a batsman is bowled out, it does not necessarily signify good fielding. So we only include catches and run out as a measure of fielding. Difference in run rates should be negative if fielding is good. Let us see if this fact is true.Figure \ref{fig:field} shows the difference between run rate between two subsequent overs are negative when good fielding leads to one or two dismissals in an over.

```{r field, fig.cap = "Distribution of lagged run rate across overs of the innings and dismissals by catch and catch and bowled."}
cricket_fielding <- cricket_season %>% 
  mutate(fielding_proxy = if_else(dismissal_kind %in% c("caught", "caught and bowled"),1,0)) %>% 
  group_by(season, match_id, batting_team, bowling_team,  inning, over) %>%
  summarise(runs_per_over = sum(total_runs),
            run_rate = sum(total_runs)/length(total_runs),
            fielding_wckts = sum(fielding_proxy)) %>%
            mutate(lagged_run_rate = c(0,diff(run_rate)))

cricket_tsibble_fld %>% 
  filter(fielding_wckts %in% c(0,1)) %>% 
  prob_plot( "fielding_wckts","over",
                               hierarchy_model,
                               response = "lagged_run_rate",
                               plot_type = "quantile",
                               quantile_prob = c(0.1, 0.25, 0.5, 0.75, 0.9),
           overlay = TRUE)  + ggtitle("")

Q4: Is run rate set to reduce in overs where number of dot balls are more/number of wickets are more?

A dot ball is a delivery bowled without any runs scored off it. The number of dot balls is reflective of the quality of bowling in the game. Run rate of an over should ideally decrease if the number of dot balls increase. However, what is the effect of dot balls in previous over on run rate in the subsequent over. Will players  batsman likely to go for big shots because they couldn't score good runs in the previous over? Or they should play consistently and avoid scoring high?

Figure \ref{fig:dot} shows that for any over, increase in dot balls lead to decrease in run rate.

```{r crickread}
cricket_tsibble <- gravitas::cricket %>%
  mutate(data_index = row_number()) %>%
  as_tsibble(index = data_index)
```

```{r exdotplot, echo=FALSE, eval = FALSE}
library(gravitas)
library(tsibble)

cricket_bowling <- cricket_season %>% 
  mutate(dot_ball_proxy = if_else(total_runs==0,1,0)) %>% 
  group_by(season, match_id, batting_team, bowling_team,  inning, over) %>%
  summarise(runs_per_over = sum(total_runs),
            run_rate = sum(total_runs)/length(total_runs),
            dot_balls = sum(dot_ball_proxy)) %>%
            mutate(diff_run_rate = c(0,diff(run_rate)))

cricket_tsibble_bow %>% gran_obs("over", "dot_balls")

cricket_tsibble_bow %>% 
  filter(dot_balls %in% c(0, 1, 2))%>% 
  prob_plot("dot_balls",
            "over",
            hierarchy_model,
            response = "diff_run_rate",
            plot_type = "quantile",
            quantile_prob = c(0.1, 0.25, 0.5, 0.75, 0.9))

```{r exwicketplot}

cricket_tsibble %>% 
  filter(dot_balls %in% c(0, 1, 2)) %>%
  mutate(lag_dot = lag(dot_balls)) %>% 
  filter(lag_dot %in% c(0, 1, 2)) %>%
  prob_plot("lag_dot",
            "over",
            hierarchy_model,
            response = "runs_per_over",
            plot_type = "quantile",
            quantile_prob = c(0.1, 0.25, 0.5, 0.75, 0.9))

cricket_tsibble %>% 
  filter(wicket %in% c(0, 1)) %>%
  mutate(lag_wicket = lag(wicket)) %>% 
  filter(lag_wicket %in% c(0, 1)) %>%
  prob_plot("lag_wicket",
            "over",
            hierarchy_model,
            response = "runs_per_over",
            plot_type = "quantile",
            quantile_prob = c(0.1, 0.25, 0.5, 0.75, 0.9))

```{r write-bib}
write_bib(c("ggplot2", "tidyverse", "sugrrants", "geofacet", "shiny", "wanderer4melb"), 'packages.bib')
```

@@

<!--   cricket_tsibble %>% prob_plot("inning_match", "over_inning", -->
<!--                                hierarchy_model, -->
<!--                                response = "runs_per_over", -->
<!--                                plot_type = "quantile", -->
<!--                                quantile_prob = c(0.1, 0.25, 0.5, 0.75, 0.9)) + -->
<!--     theme( # remove the vertical grid lines -->
<!--            panel.grid.major.x = element_blank() , -->
<!--            # explicitly set the horizontal lines (or they will disappear too) -->
<!--            panel.grid.major.y = element_line( size=.1, color="black" )  -->
<!--     ) -->

@@

<!-- cricket_tsibble_all %>% -->
<!--   granplot( -->
<!--     gran1 = "inning_match", -->
<!--     gran2 = "over_inning", -->
<!--     hierarchy_model, -->
<!--     response = "total_runs", -->
<!--     plot_type = "quantile" -->
<!--   ) -->
<!-- ``` --
```

>>>>

The application is not only restricted to temporal data. We provide an example of cricket to illustrate how this can be generalized in other applications. The Indian Premier League (IPL) is a professional Twenty20 cricket league in India contested by eight teams representing eight different cities in India. With eight teams, each team plays each other twice in a home-and-away round-robin format in the league phase. In a Twenty20 game the two teams have a single innings each, which is restricted to a maximum of 20 overs. Hence, in this format of cricket, a match will consist of 2 innings, an innings will consist of 20 overs, an over will consist of 6 balls with some exceptions.

@@

<!-- ```{r hierarchy, echo=FALSE} -->
<!-- hierarchy_model <- tibble::tibble(linear gran = c("ball", "over", "inning", "match"),  -->
<!--                                   convert_fct = c(6, 20, 2, 1)) -->
<!-- knitr::kable(hierarchy_model, caption = " A hierarchy table for T20 cricket") -->
<!-- ``` -->

@@

```{r hierarchy2}
library(gravitas)
library(tibble)
hierarchy_model <- tibble::tibble(
  G = c("over", "inning", "match"),
  k = c(20, 2, 1))

@@

First, we look at the distribution of runs per over across over of the innings and seasons in \autoref{fig:seas-over-inning}. The distribution of runs per over has not significantly changed from 2008 to 2016. There is no clear pattern/trend that runs per over is increasing or decreasing across seasons. Hence, we work with subsets of seasons to answer some of the questions:

Mumbai Indians(MI) and Chennai Super kings(CSK) are considered one of the best teams in IPL with multiple winning titles and always appearing in final 4 from 2010 to 2015. It would be interesting to take their example in order to dive deeper into the first question.

@@

@@

```{r seas-over-inning, fig.cap = "Quantile plot of runs per over across overs of different seasons. There is no pattern on increase or decrease of runs across overs for seasons."}
  cricket_tsibble_all %>% 
  prob_plot("over", 
            "season",
            hierarchy_model,
            response = "runs_per_over",
            plot_type = "quantile",
            quantile_prob = c(0.25, 0.5, 0.75), 
            symmetric = FALSE)  +
     theme( # remove the vertical grid lines
            panel.grid.major.x = element_blank() ,
            # explicitly set the horizontal lines (or they will disappear too)
            panel.grid.major.y = element_blank()) + scale_x_discrete(breaks = seq(2008, 2016, 3))
```

@@

cricket_tsibble %>%
  filter(batting_team %in% c("Mumbai Indians",
                             "Chennai Super Kings"))%>%
  prob_plot("inning", 
            "over",
  response = "runs_per_over",
  hierarchy_model,
  plot_type = "lv")  + ggtitle("")
```

Q2: Is run rate set to reduce in subsequent over if fielding/bowling is good in the previous over? Between fielding and bowling which penalizes runs in the subsequent over more?

For establishing that the fielder fielded well in a particular over, we can see how many catches and run outs were made in that particular over. If a batsman is bowled out, it does not necessarily signify good fielding. So we only include catches and run out as a measure of fielding. Difference in runs across over should be negative if good fielding has an impact on the runs scored in the subsequent overs. Let us see if this fact is true. Figure \ref{fig:field} shows the difference between run rate between two subsequent overs are negative when good fielding leads to one or two dismissals in an over, implying good fielding in one over has indeed an impact on runs scored in the subsequent overs.

```{r field, fig.cap = "Distribution of lagged run rate across overs of the innings and dismissals by catch and catch and bowled. This shows good fielding in one over leads to lower runs in the subsequent overs for at least 50 percentiles of times."}

cricket_fielding <- cricket_season %>% 
  mutate(fielding_proxy = if_else(dismissal_kind %in% c("caught", "caught and bowled"),1,0)) %>% 
  group_by(season, match_id, batting_team, bowling_team,  inning, over) %>%
  summarise(runs_per_over = sum(total_runs),
            run_rate = sum(total_runs)/length(total_runs),
            fielding_wckts = sum(fielding_proxy)) %>%
            mutate(diff_run_rate = c(0,diff(run_rate)))

cricket_tsibble_fld %>% 
  filter(fielding_wckts %in% c(0,1)) %>% 
  prob_plot( "fielding_wckts","over",
                               hierarchy_model,
                               response = "diff_run_rate",
                               plot_type = "quantile",
                               quantile_prob = c(0.1, 0.25, 0.5, 0.75, 0.9),
           overlay = TRUE)  + ggtitle("")

<!-- Q2: Amongst good fielding and bowling - which affect the runs of the subsequent overs more? -->

Q3: Is runs set to reduce in the next over for dot balls in this over?

A dot ball is a delivery bowled without any runs scored off it. The number of dot balls is reflective of the quality of bowling in the game. Run rate of an over should ideally decrease if the number of dot balls increase. However, what is the effect of dot balls on runs scored in the subsequent over. Will players  batsman likely to go for big shots because they couldn't score good runs in the previous over? Or they should play consistently and avoid scoring high? From figure \autoref{exdot} shows the quantile plot of runs across overs for none, one or two dot balls per over in the facets. Each of the quantiles decrease across facets. With two dot balls per over, at least 50% of the times, run rates decrease in the subsequent over as observed through the negative values for 10th, 25th and 50th percentile quantiles.

```{r crickread, ech0=FALSE}
cricket_tsibble <- gravitas::cricket %>%
  mutate(data_index = row_number()) %>%
  as_tsibble(index = data_index)
```

```{r exdot, fig.cap="10th, 20th, 30th, 40th and 50th quantiles of runs per over are drawn  across overs of the innings with no (facet 1), one (facet 2) and two (facet 3) dot balls per over. For all three cases, difference in runs between subsequent overs decrease."}

cricket_bowling <- cricket_season %>% 
  mutate(dot_ball_proxy = if_else(total_runs==0,1,0),
         wicket_proxy = if_else(is.na(dismissal_kind),0,1)) %>% 
  group_by(season, match_id, batting_team, bowling_team,  inning, over) %>%
  summarise(dot_balls = sum(dot_ball_proxy),
            wickets = sum(wicket_proxy), 
            total_runs = sum(total_runs)) %>%
            mutate(diff_run_rate = c(0,diff(total_runs)))

cricket_tsibble_bow %>%
  filter(dot_balls %in% c(0, 1, 2)) %>%
  prob_plot("dot_balls",
            "over",
            hierarchy_model,
            response = "diff_run_rate",
            plot_type = "quantile",
            quantile_prob = c(0.1, 0.2, 0.3, 0.4, 0.5),
            symmetric = FALSE)  + ggtitle("")

<!-- # ```{r write-bib} -->
<!-- # write_bib(c("ggplot2", "tidyverse", "sugrrants", "geofacet", "shiny", "wanderer4melb"), 'packages.bib') -->
<!-- # ``` -->

<!-- library(gravitas) -->
<!-- library(tsibble) -->

<!--  cricket_bowling <- cricket_season %>% -->
<!--    mutate(dot_ball_proxy = if_else(total_runs==0,1,0), -->
<!--           wicket_proxy = if_else(is.na(dismissal_kind),0,1)) %>% -->
<!--   group_by(season, match_id, batting_team, bowling_team,  inning, over) %>% -->
<!--    summarise(dot_balls = if_else(sum(dot_ball_proxy)>0, "yes","no"), -->
<!--              wickets = if_else(sum(wicket_proxy)>0, "yes","no"),  -->
<!--             total_runs = sum(total_runs)) %>% -->
<!--             mutate(diff_run_rate = c(0,diff(total_runs))) -->
<!--   -->
<!--   -->
<!--  cricket_tsibble_bow <- cricket_bowling %>%  -->
<!--    ungroup() %>%  -->
<!--    mutate(data_index = row_number()) %>%  -->
<!--    as_tsibble(index = data_index) -->

<!-- cricket_tsibble_bow %>% gran_obs("over", "dot_balls") -->

<!-- # cricket_tsibble_bow %>%  -->
<!-- #   #filter(dot_balls %in% c(0, 1, 2))%>%  -->
<!-- #   prob_plot("dot_balls", -->
<!-- #             "over", -->
<!-- #             hierarchy_model, -->
<!-- #             response = "diff_run_rate", -->
<!-- #             plot_type = "quantile", -->
<!-- #             quantile_prob = c(0.1, 0.25, 0.5, 0.75, 0.9), -->
<!-- #             symmetric = FALSE) -->

<!-- cricket_tsibble_bow %>% gran_obs("over", "wickets") -->

<!-- cricket_tsibble_bow %>%  -->
<!--   #filter(wickets %in% c(0, 1))%>%  -->
<!--   gather("proxies", "yes_no", -c(season, -->
<!--                                      match_id, -->
<!--                                      batting_team, -->
<!--                                      bowling_team, -->
<!--                                      inning,     -->
<!--                                      over, -->
<!--                                      diff_run_rate, -->
<!--                                      data_index)) %>%  -->
<!--   filter(yes_no %in% c("no")) %>%  -->
<!--   prob_plot("proxies", -->
<!--             "over", -->
<!--             hierarchy_model, -->
<!--             response = "diff_run_rate", -->
<!--             plot_type = "boxplot", -->
<!--             symmetric = FALSE, -->
<!--             quantile_prob = c(0.1, 0.25, 0.5, 0.75, 0.9))+ ggtitle("") -->
<!-- ``` -->

<!-- ```{r proxies-yes} -->
<!-- cricket_tsibble_bow %>%  -->
<!--   #filter(wickets %in% c(0, 1))%>%  -->
<!--   gather("proxies", "yes_no", -c(season, -->
<!--                                      match_id, -->
<!--                                      batting_team, -->
<!--                                      bowling_team, -->
<!--                                      inning,     -->
<!--                                      over, -->
<!--                                      diff_run_rate, -->
<!--                                      data_index)) %>%  -->
<!--   filter(yes_no %in% c("yes")) %>%  -->
<!--   prob_plot("proxies", -->
<!--             "over", -->
<!--             hierarchy_model, -->
<!--             response = "diff_run_rate", -->
<!--             plot_type = "lv", -->
<!--             symmetric = FALSE, -->
<!--             quantile_prob = c(0.1, 0.25, 0.5, 0.75, 0.9))+ ggtitle("") -->

<!-- ```{r exwicketplot, echo=FALSE} -->
<!-- #  -->
<!-- # cricket_tsibble %>%  -->
<!-- #   filter(dot_balls %in% c(0, 1, 2)) %>% -->
<!-- #   mutate(lag_dot = lag(dot_balls)) %>%  -->
<!-- #   filter(lag_dot %in% c(0, 1, 2)) %>% -->
<!-- #   prob_plot("lag_dot", -->
<!-- #             "over", -->
<!-- #             hierarchy_model, -->
<!-- #             response = "runs_per_over", -->
<!-- #             plot_type = "quantile", -->
<!-- #             quantile_prob = c(0.1, 0.25, 0.5, 0.75, 0.9)) -->

<!-- cricket_tsibble %>%  -->
<!--   filter(wicket %in% c(0, 1)) %>% -->
<!--   mutate(lag_wicket = lag(wicket)) %>%  -->
<!--   filter(lag_wicket %in% c(0, 1)) %>% -->
<!--   prob_plot("lag_wicket", -->
<!--             "over", -->
<!--             hierarchy_model, -->
<!--             response = "runs_per_over", -->
<!--             plot_type = "quantile", -->
<!--             quantile_prob = c(0.1, 0.25, 0.5, 0.75, 0.9)) -->

@@

<!-- r lv, fig.cap = "Letter value plot of run rate across overs of an inning and innings of the match for (top) Mumbai Indians and (bottom) Chennai Super Kings."} -->
# 
# 
# 
# MI <- cricket_tsibble_all %>%
#   filter(season %in% c(2010:2015), batting_team=="Mumbai Indians")
# 
# 
# CSK <- cricket_tsibble_all %>%
#   filter(season %in% c(2010:2015),
#          batting_team=="Kolkata Knight Riders",
#          inning %in% c(1,2))
# 
# 
#  MI %>%
#    gran_obs("inning", "over", hierarchy_model) %>%
#    knitr::kable(format = "markdown", caption = "Cross tabulation of Innings and Overs for Mumbai Indians: 2010-2015")
# 
#  CSK %>%
#    gran_obs("inning", "over", hierarchy_model) %>%
#    knitr::kable(format = "markdown", caption = "Cross tabulation of Innings and Overs for Mumbai Indians: 2010-2015")
# 
# MI %>%
#   prob_plot("inning", "over",
#                                hierarchy_model,
#                                response = "run_rate",
#                                plot_type = "lv",
#                                quantile_prob = c(0.1, 0.25, 0.5, 0.75, 0.9))  + ggtitle("")
# 
# 
# CSK %>%
#   prob_plot("inning", "over",
#                                hierarchy_model,
#                                response = "run_rate",
#                                plot_type = "lv",
#                                quantile_prob = c(0.1, 0.25, 0.5, 0.75, 0.9)) + ggtitle("")
# 
# cricket_tsibble_all %>% distinct(batting_team)

@@

<!--   cricket_tsibble %>% prob_plot("inning_match", "over_inning", -->
<!--                                hierarchy_model, -->
<!--                                response = "runs_per_over", -->
<!--                                plot_type = "quantile", -->
<!--                                quantile_prob = c(0.1, 0.25, 0.5, 0.75, 0.9)) + -->
<!--     theme( # remove the vertical grid lines -->
<!--            panel.grid.major.x = element_blank() , -->
<!--            # explicitly set the horizontal lines (or they will disappear too) -->
<!--            panel.grid.major.y = element_line( size=.1, color="black" )  -->
<!--     )

@@

<!-- cricket_tsibble_all %>% -->
<!--   granplot( -->
<!--     gran1 = "inning_match", -->
<!--     gran2 = "over_inning", -->
<!--     hierarchy_model, -->
<!--     response = "total_runs", -->
<!--     plot_type = "quantile" -->
<!--   ) -->
